Extreme climatic events with asymmetric tail-dependence affect population-extinction risk

### 2019-07-25 MainText.Rmd
edit intro

# Introduction \label{intro}

@@

<!--How copula could be relevent in estimating extreme event's 
influence on extinction risk : introduce Taildep. concept-->
A recent study [@Ghosh_copula] showed the importance, causes 
and consequences of copula structures in associations between ecologically
important variables and demonstrated one example of density-independent metapopulation model where 
copula was used to estimate extinction risk. Copula, a rank based measure of association, is much more informative 
than linear correlation measure and a widely applied tool in dependence 
modeling [@joe2014_dependence; @nelsen2006_copula]. We here introduce the concept of copula in a
manner that follows @Ghosh_copula. Copulas separate the information content of a bivariate dataset,
$(x_n, y_n)$ for $n = 1 \ldots N$, into two non-overlapping parts: the information in the marginal 
distributions (which is not about the association between the variables) and the rest of 
the information (which is solely about the association). The association information is revealed by 
the copula plot ($v_n$ against $u_n$) of the raw variables ($y_n$ vs. $x_n$) in Fig...., where 
$u_n$ is the rank of $x_n$ in the set ${x_1, x_2, \ldots , x_n }$, divided by $N + 1$; 
and $v_n$ is the rank of $y_n$ in the set ${y_1, y_2, \ldots , y_n }$, also divided
by $N + 1$. Here the rank of the smallest element of each sets is considered to be 1 and the rank of
the largest element is $N$. We refer to the $u_n$ and $v_n$ as normalized ranks of the $x_n$ and $y_n$. 
Ranking makes the marginal distributions uniform, isolating only the information on 
association between the variables. Here, we will extend the application of copula 
in estimating extinction risk as introduced by [@Ghosh_copula] on why the potential use of copula
in modelling extinction risk goes beyond the density-independent model 
and even applicable for a simple age-structured population where its vital rates are influenced by ECEs.

<!--tail dep. of copula & extinction risk-->
A major benefit of the copula approach is that it can effectively capture the tail dependence. 
Tail dependence, already exemplified in Fig..... (add Fig.), is association between the more extreme 
values of distributions. For two positively associated variables, left-tail 
(respectively, right-tail) dependence is stronger association between
values in the left or lower portions, i. e. for smaller values 
(respectively, the right or upper portions, i.e. for higher values) 
of the two distributions. Stronger dependence in either tail gives rise to asymmetric copula 
structure for the joint distribution of two variables. 
For example, extreme cold winter during growing season
might reduce the adult reproduction rate as well as larval survival (i.e. stronger 
tail dependence between vital rates when temperature is too cold than normal), which 
could lead to population crash and thereby inflate the extinction risk. Similarly, in context of 
density-dependent metapopulation models, if extreme environmental fluctuations across sites 
appear to be as synchronous (left tail dependent or right tail dependent between patches 
with same correlation as in Fig.....) then it could wipe out some metapopulation patches 
during unfavorable years unless they are rescued by the nearby populations. This scenario 
could be more complex depending 
on the dynamics of the model (see \nameref{results}). Though tail-dependence 
property of a copula has been used for risk-analysis in other fields like finance, 
hydrology etc. [@alexander2008market; @li2013; @chen2011flood; @Goswami2018], but to our knowledge
copula have never been used in estimating the effect of ECEs through extinction risk of age structured 
model or density-dependent metapopulations.

<!--our objective ?-->
In this paper, we will explore the potential use of asymmetric tail-dependence in a copula 
for understanding the extinction risk of populations towards extreme climatic event in ecology
by addressing the following questions: (Q1) Does nature of tail dependence 
(left tail or right tail) in vital rates
of a single-patch age-structured population model affect extinction risk of that population differently?
We hypothesized population might have higher extinction risk if vital rates are left-tail dependent.
(Q2) How does asymmetric tail-dependent noises across multiple patches influence extinction 
risk of a nonlinear metapopulation model with density dependence? Our speculation is that the general finding 
from density-independent metapopulation model [@Ghosh_copula]
(higher extinction risk if noises across patches were 
left-tail dependent than the one if noises had right-tail dependence) holds true only in
the under-compensatory regime, but should be reverse if the model shows over-compensatory dynamics. 
(Q3) We further hypothesize that when there is no dispersal, the above-said switching should occur
near the critical parameter value where the deterministic model shows transition from 
under-compensatory to over-compensatory dynamics.

<!--Tuljapurkar estimation to be explained in Discussion-->

# Methods \label{methods}

<!--stochastic matrix model : age structured-->
To investigate (Q1) from \nameref{intro} that how copula structure in noisy 
environment affect the extinction risk of
populations in a single patch age-structured model,
we considered a basic stochastic age-structured model (Eq. \ref{eq_age_str}) 
with two age classes (egg $E$ and adult $A$). We assumed
the population was semelperous, i.e., adult ($A$) went through the
reproductive episode only once before death. Eggs and adults in the 
next generation depend upon the fecundity of adult ($f_a \geq 0$) 
and survival of eggs ($0 \leq s_e \leq 1$) from present generation, respectively.
The environmental stochasticity in the model was embedded in 
the fluctuating vital rates ($f_a$ and $s_e$). We considered the fecundity ($f_a$)
and survival ($s_e$) rates were random draw from a gamma ($\Gamma(k,\theta)$) and
a beta ($\mathcal{B}(\alpha,\beta)$) distribution, respectively. These distributions 
could be generated from iid noises showing specific tail-dependence (left or right tail)
between fecundity and survival rates. Eq. \ref{eq_vitalrates} represent the stochasticity 
in the vital rates. We included the density dependence effect in one of the vital rate,
namely, fecundity ($f_a$). If $\Omega = \infty$ in Eq. \ref{eq_vitalrates}, 
the projection matrix would be independent
of populations from any age-class and the model was then called to be "linear", whereas
its "nonlinear" version had finite $\Omega$.

@@

<!--test hypothesis from Q2 : ext risk with density-dependence metapopulation model with r from different regime-->
To test our hypothesis from (Q2) of \nameref{intro} for each metapopulation model,
we estimated extinction risk for a wide range of parameter $r$, 
which covered the undercompensatory and overcompensatory regime. We 
chose equispaced values of $r$ from the 95\% quantiles 
of possible range ($r_{min},r_{bf}$), where $r_{min}$ was the minimum value of $r$ 
for which the model equilibrium $P_e$ had always a finite, positive initial 
population (for details, see Appendix \ref{SM-model_stability}). For each chosen $r$ from each 
model where metapopulation could disperse locally with varying dispersal 
rate $d= 0,0.1,0.2,..0.5$, we calculated extinction risk with left 
tail dependent and right tail dependent environmental noise 
in different simulations, respectively. We also repeated the analysis 
for global dispersal for each metapopulation model. Script 
[`mydispersal.R`](https://github.com/sghosh89/ERC/blob/master/mydispersal.R) was used for this 
analysis.

@@

# Results \label{results}

@@

We found density-dependence metapopulation models (with $n=5$ patches) in noisy 
environment showed similar patterns for extinction risk as density-independent 
metapopulation model showed in earlier studies [@Ghosh_copula]. For each density-dependent 
model listed in Table \ref{tab_model_summary}, we observed higher (or lower) extinction risk for 
the metapopulations when noises across multiple patches were left- (or right) tail dependent.
But that observation remained valid when the model parameter, usually growth rate $(r)$, of each model was 
sufficiently low so that under-compensatory dynamics (monotonous approach towards
its stable equilibrium) for the deterministic analog of the model was ensured. When the model 
was in over-compensatory regime (oscillatory approach towards stable equilibrium) for higher 
value of $r$, we observed right (or left) tail dependent noise caused higher (or lower) extinction 
risk for the metapopulations. Fig. \ref{fig_extrisk_allmodel_localdisp} shows variation of extinction risk 
against local dispersal rate ($d$) for six density dependent metapopulation models 
where growth rate $r$ was taken either from under-compensatory or from over-compensatory 
regime. We also got similarly consisent results with 
our prediction (see Q2 from \nameref{intro}) for global 
dispersal (Fig. \ref{SM-fig_extrisk_allmodel_globaldisp}). 
Figs. (\ref{SM-fig_ricker_risk_local_disp_scl_1} - \ref{SM-fig_risk_local_disp_abrewer_r_vary_scl_0.6}) 
shows how extinction risk changed against local dispersal rate for a whole range
of $r$, whereas Figs. (\ref{SM-fig_ricker_risk_global_disp_scl_1} - 
\ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6}) 
represents the same but for the case when global dispersal took place between patches. 
Both case showed the signature of 
"rescue-effect" (i.e. by lowering extinction risk) when there was 
a finite dispersal compared to the one with 
$d=0$.

<!--Q3 : metapopulation results : switching near rc for d=0, Fig. 3-->
Following Figs. (\ref{SM-fig_ricker_risk_local_disp_scl_1} - 
\ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6}) we hypothesized that the switching 
between the extinction risk generated by left-tail dependent and right-tail dependent noise 
occurred near the critical parameter $r_c$ (see Q3). For simplicity we only consider 
$d=0$ case, i.e., when there was no dispersal between patches (see \nameref{methods}). 
Fig. \ref{fig_extrisk_d0} shows higher extinction risk with left-tail dependent noise 
for $r<r_c$ and with right-tail dependent noise for $r>r_c$. $r_c$ was the vertical
line in Fig. \ref{fig_extrisk_d0} indicating the value of $r$ for each model where 
their deterministic analog switches from under- to over-compensatory dynamics. 
Though the switching visually appered far from the theoretical $r_c$ value for Verhulst 
and Austin-Brewer model, they were consistent with our hypothesis as they had comperatively 
smaller range of $r$ than the other models.

@@

>>>>

# Introduction\label{Introduction}

@@

<!--How copula could be relevent in estimating extreme event's 
influence on extinction risk : introduce Taildep. concept-->
A recent study [@Ghosh_copula] showed the importance, causes 
and consequences of copula structures in associations between ecologically
important variables and demonstrated a basic example where 
copula was used to model extinction risk of a density-independent 
metapopulation. Present manuscript is an extension for the application of copula 
in estimating the extinction risk as introduced by [@Ghosh_copula]. We will show 
why the potential use of copula
in modelling extinction risk goes beyond the density-independent model 
and even applicable for a simple age-structured population where its vital rates are influenced by ECEs. 
Following we will introduce the concept of copula in a manner that follows @Ghosh_copula.

<!--brief intro for copula-->
Copula, a rank based measure of association, is much more informative 
than linear correlation measure and a widely applied tool in dependence 
modeling [@joe2014_dependence; @nelsen2006_copula]. Copulas separate the information content of a bivariate dataset,
$(x_n, y_n)$ for $n = 1 \ldots N$, into two non-overlapping parts: the information in the marginal 
distributions (which is not about the association between the variables) and the rest of 
the information (which is solely about the association). The association information is revealed by 
the copula plot ($v_n$ against $u_n$) of the raw variables ($y_n$ vs. $x_n$) in Fig...., where 
$u_n$ is the rank of $x_n$ in the set ${x_1, x_2, \ldots , x_n }$, divided by $N + 1$; 
and $v_n$ is the rank of $y_n$ in the set ${y_1, y_2, \ldots , y_n }$, also divided
by $N + 1$. Here the rank of the smallest element of each sets is considered to be 1 and the rank of
the largest element is $N$. We refer to the $u_n$ and $v_n$ as normalized ranks of the $x_n$ and $y_n$. 
Ranking makes the marginal distributions uniform, isolating only the information on 
association between the variables.

<!--tail dep. of copula & extinction risk-->
A major benefit of the copula approach is that it can effectively capture the tail dependence. 
Tail dependence, already exemplified in Fig..... (add Fig.), is association between the more extreme 
values of distributions. For two positively associated variables, left-tail 
(respectively, right-tail) dependence is stronger association between
values in the left or lower portions, i. e. for smaller values 
(respectively, the right or upper portions, i.e. for higher values) 
of the two distributions. Stronger dependence in either tail gives rise to asymmetric copula 
structure for the joint distribution of two variables. 
For example, extreme cold winter during growing season
might reduce the adult reproduction rate as well as larval survival (i.e. stronger 
tail dependence between vital rates when temperature is too cold than normal), which 
could lead to population crash and thereby inflate the extinction risk. Similarly, in context of 
density-dependent metapopulation models, if extreme environmental fluctuations across sites 
appear to be as synchronous (left tail dependent or right tail dependent between patches 
with same correlation as in Fig.....) then it could wipe out some metapopulation patches 
during unfavorable years unless they are rescued by the nearby populations. This scenario 
could be more complex depending 
on the dynamics of the model (see \nameref{Results}). Though tail-dependence 
property of a copula has been used for risk-analysis in other fields like finance, 
hydrology etc. [@alexander2008market; @li2013; @chen2011flood; @Goswami2018], but to our knowledge
copula have never been used in estimating the effect of ECEs through extinction risk of age structured 
model or density-dependent metapopulations.

<!--our objective ?-->
In this paper, we will explore the potential use of asymmetric tail-dependence in a copula 
for understanding the extinction risk of populations towards extreme climatic event
by addressing the following questions: (Q1) Does nature of tail dependence 
(left tail or right tail) in vital rates
of a single-patch age-structured population model affect extinction risk of that population differently?
We hypothesized population might have higher extinction risk if vital rates are left-tail dependent.
(Q2) How does asymmetric noise copula (tail dependence across multiple patches) influence extinction 
risk of a density-dependent metapopulation model? Our speculation is that the general finding 
from density-independent metapopulation model [@Ghosh_copula]
(higher extinction risk if noises across patches were 
left-tail dependent than the one if noises had right-tail dependence) would hold true only in
the under-compensatory regime, but should be reverse if the model shows over-compensatory dynamics. 
(Q3) We further hypothesize that when there is no dispersal, the above-said switching should occur
near the critical parameter value where the deterministic model shows transition from 
under-compensatory to over-compensatory dynamics.

# Methods\label{Methods}

<!--stochastic matrix model : age structured-->
To investigate (Q1) from \nameref{Introduction} that how copula structure in noisy 
environment affect the extinction risk of
populations in a single patch age-structured model,
we considered a basic stochastic age-structured model (Eq. \ref{eq_age_str}) 
with two age classes (egg $E$ and adult $A$). We assumed
the population was semelperous, i.e., adult ($A$) went through the
reproductive episode only once before death. Eggs and adults in the 
next generation depend upon the fecundity of adult ($f_a \geq 0$) 
and survival of eggs ($0 \leq s_e \leq 1$) from present generation, respectively.
The environmental stochasticity in the model was embedded in 
the fluctuating vital rates ($f_a$ and $s_e$). We considered the fecundity ($f_a$)
and survival ($s_e$) rates were random draw from a gamma ($\Gamma(k,\theta)$) and
a beta ($\mathcal{B}(\alpha,\beta)$) distribution, respectively. These distributions 
could be generated from iid noises showing specific tail-dependence (left or right tail)
between fecundity and survival rates. Eq. \ref{eq_vitalrates} represent the stochasticity 
in the vital rates. We included the density dependence effect in one of the vital rate,
namely, fecundity ($f_a$). If $\Omega = \infty$ in Eq. \ref{eq_vitalrates}, 
the projection matrix would be independent
of populations from any age-class and the model was then called to be "linear", whereas
its "nonlinear" version had finite $\Omega$.

@@

<!--test hypothesis from Q2 : ext risk with density-dependence metapopulation model with r from different regime-->
To test our hypothesis from (Q2) of \nameref{Introduction} for each metapopulation model,
we estimated extinction risk for a wide range of parameter $r$, 
which covered the undercompensatory and overcompensatory regime. We 
chose equispaced values of $r$ from the 95\% quantiles 
of possible range ($r_{min},r_{bf}$), where $r_{min}$ was the minimum value of $r$ 
for which the model equilibrium $P_e$ had always a finite, positive initial 
population (for details, see Appendix \ref{SM-model_stability}). For each chosen $r$ from each 
model where metapopulation could disperse locally with varying dispersal 
rate $d= 0,0.1,0.2,..0.5$, we calculated extinction risk with left 
tail dependent and right tail dependent environmental noise 
in different simulations, respectively. We also repeated the analysis 
for global dispersal for each metapopulation model. Script 
[`mydispersal.R`](https://github.com/sghosh89/ERC/blob/master/mydispersal.R) was used for this 
analysis.

@@

# Results\label{Results}

@@

We found density-dependence metapopulation models (with $n=5$ patches) in noisy 
environment showed similar patterns for extinction risk as density-independent 
metapopulation model showed in earlier studies [@Ghosh_copula]. For each density-dependent 
model listed in Table \ref{tab_model_summary}, we observed higher (or lower) extinction risk for 
the metapopulations when noises across multiple patches were left- (or right) tail dependent.
But that observation remained valid when the model parameter, usually growth rate $(r)$, of each model was 
sufficiently low so that under-compensatory dynamics (monotonous approach towards
its stable equilibrium) for the deterministic analog of the model was ensured. When the model 
was in over-compensatory regime (oscillatory approach towards stable equilibrium) for higher 
value of $r$, we observed right (or left) tail dependent noise caused higher (or lower) extinction 
risk for the metapopulations. Fig. \ref{fig_extrisk_allmodel_localdisp} shows variation of extinction risk 
against local dispersal rate ($d$) for six density dependent metapopulation models 
where growth rate $r$ was taken either from under-compensatory or from over-compensatory 
regime. We also got similarly consisent results with 
our prediction (see Q2 from \nameref{Introduction}) for global 
dispersal (Fig. \ref{SM-fig_extrisk_allmodel_globaldisp}). 
Figs. (\ref{SM-fig_ricker_risk_local_disp_scl_1} - \ref{SM-fig_risk_local_disp_abrewer_r_vary_scl_0.6}) 
shows how extinction risk changed against local dispersal rate for a whole range
of $r$, whereas Figs. (\ref{SM-fig_ricker_risk_global_disp_scl_1} - 
\ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6}) 
represents the same but for the case when global dispersal took place between patches. 
Both case showed the signature of 
"rescue-effect" (i.e. by lowering extinction risk) when there was 
a finite dispersal compared to the one with 
$d=0$.

<!--Q3 : metapopulation results : switching near rc for d=0, Fig. 3-->
Following Figs. (\ref{SM-fig_ricker_risk_local_disp_scl_1} - 
\ref{SM-fig_risk_global_disp_abrewer_r_vary_scl_0.6}) we hypothesized that the switching 
between the extinction risk generated by left-tail dependent and right-tail dependent noise 
occurred near the critical parameter $r_c$ (see Q3). For simplicity we only consider 
$d=0$ case, i.e., when there was no dispersal between patches (see \nameref{Methods}). 
Fig. \ref{fig_extrisk_d0} shows higher extinction risk with left-tail dependent noise 
for $r<r_c$ and with right-tail dependent noise for $r>r_c$. $r_c$ was the vertical
line in Fig. \ref{fig_extrisk_d0} indicating the value of $r$ for each model where 
their deterministic analog switches from under- to over-compensatory dynamics. 
Though the switching visually appered far from the theoretical $r_c$ value for Verhulst 
and Austin-Brewer model, they were consistent with our hypothesis as they had comperatively 
smaller range of $r$ than the other models.

@@

<!--summary results-->

<!--Tuljapurkar approximation : linear version of age-structured model in stochastic environments-->

<!--future study: more complex stage or age structured model, extinction risk how propagated through a food web-->

<!--future study: worthy for PVA, IUCN red list-->

<!--match-mismatch hypothesis leads to extinction? our phenology paper in prep.-->
